Bayesian model in JAGS

### 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
updates

There are 5 main statistical methods to estimate the hospitalization/mortality attributable to respiratory virus infection. In Dan's class, Public Health Surveillance, he teaches all four methods in details. Here, we will only give a brief introduction to each of the method and provide the link to the initial publications.
\
\
1. **Serfling regression**

@@

1. **Serfling regression** \
  This type of model identifies the epidemic season and establish a epidemic threshold (seasonal baseline) using historical data. After predicting the expected diseases baseline, observed diseases above the baseline (epidemic threshold) during the epidemic season will be attributable to the viral infection. The basic Serfling regression is given by: $Y_t = \mu+ bt+\sum \alpha_i cos \theta+ \sum \beta_i sin \theta$ where $Y_t$ is the expected hospitalizations at time t. $\mu$ is the baseline. $bt$ captures time trend. $\sum \alpha_i cos \theta+ \sum \beta_i sin \theta$ are for the seasonal variations. Serfling regression is a linear model initially using ordinary least square to fit. [\color{blue} Click to see an R code example](https://kmcconeghy.github.io/flumodelr/articles/04-serfling.html)
  \
  \
2. **Periseason differences**

>>>>

There are 5 main statistical methods to estimate the hospitalization/mortality attributable to respiratory virus infection. In Dan's class, Public Health Surveillance, he teaches all four methods in details. Here, we will only give a brief introduction to each of the method and provide the link to the initial publications.
\
\

1. **Serfling regression**

@@

1.**Serfling regression** \
  This type of model identifies the epidemic season and establish a epidemic threshold (seasonal baseline) using historical data. After predicting the expected diseases baseline, observed diseases above the baseline (epidemic threshold) during the epidemic season will be attributable to the viral infection. The basic Serfling regression is given by: $Y_t = \mu+ bt+\sum \alpha_i cos \theta+ \sum \beta_i sin \theta$ where $Y_t$ is the expected hospitalizations at time t. $\mu$ is the baseline. $bt$ captures time trend. $\sum \alpha_i \cos \theta+ \sum \beta_i \sin \theta$ are for the seasonal variations. Serfling regression is a linear model initially using ordinary least square to fit. [\color{blue} Click to see an R code example](https://kmcconeghy.github.io/flumodelr/articles/04-serfling.html)
  \
  \
2.**Periseason differences**\
 This method calculate the hospitalizations attributable to influenza based on the differences in the hospitalization rates when respiratory virus was circulating and the hospitalization rates when there was no respiratory virus circulation in the community.
   \
  \

3.**Poisson regression with log link**
\
  Poisson regression is used to model count variables. Therefore, it can be used to model the number of hospitalizations over time. The logarithm of the expected value of the count variable is modeled as a linear combination of the predictor variables. For example, in Thompson's paper^[Thompson WW, Shay DK, Weintraub E, et al. Mortality Associated With Influenza and Respiratory Syncytial Virus in the United States. JAMA. 2003;289(2):179–186. doi:10.1001/jama.289.2.179], the poisson regression is given by: \small
  $$Y = \alpha \exp^{(\beta_0 + \beta_1t + \beta_2t^2  +  \beta_3 \sin(2\pi/52) + \beta_4 \cos(2\pi/52) + \beta_5 Flu_A(H1N1) + \beta_6 Flu_A(H3N2) + \beta_7Flu_B + \beta_8RSV)}$$ 
  In this example, $Y$ is the observed number of deaths/hospitalization in a particular week, $\alpha$ is the population size as an offset term,$\beta_0$ is the baseline hospitalization incidence, $\beta_1t + \beta_2t^2$ represents the long-term trend of hospitalizations, $\sin(2\pi/52) + \beta_4 \cos(2\pi/52)$ captures the annual seasonal variations of hospitalizations, the rest are the pathogens that can contribute to the increase of the observed hospitalizations.
